---
title: "Seurat v5 scRNA-seq pipeline"
output: html_document
---

## 00. Project Metadata

```{r project-metadata}
project_id <- "<GSEID>"

meta <- c(
  project_GEOid = "<GSEID>",
  cancer_type = "<TCGA_ABBREVIATION_OR_NA>",
  platform = "<PLATFORM_OR_NA>",
  patient_id = "<PATIENT_IDS_OR_NA>",
  sample_id = "<SAMPLE_IDS_OR_NA>",
  sample_type = "<T_N_L_OR_NA>"
)

sample_meta <- data.frame(
  sample_id = c("<GSM_OR_SAMPLE_ID>"),
  patient_id = c("NA"),
  sample_type = c("NA"),
  sra_run = c("NA"),
  platform = c("<PLATFORM_OR_NA>"),
  stringsAsFactors = FALSE
)
```

## 01. Environment Setup

```{r environment-setup}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)

require_or_stop <- function(pkg) {
  if (!requireNamespace(pkg, quietly = TRUE)) {
    stop(sprintf("Package '%s' is required. Install it before running this notebook.", pkg), call. = FALSE)
  }
}

required_packages <- c("Seurat", "qs", "Matrix", "data.table", "dplyr", "ggplot2", "DoubletFinder")
invisible(lapply(required_packages, require_or_stop))

library(Seurat)
library(qs)
library(Matrix)
library(data.table)
library(dplyr)
library(ggplot2)

# 运行前请先把工作目录替换为当前项目目录。
work_dir <- "<WORK_DIR>"
seed <- <SEED>
setwd(work_dir)
set.seed(seed)

input_dir <- file.path(work_dir, "data", "<GSEID>")
output_dir <- file.path(work_dir, "output", "<GSEID>")
fig_dir <- file.path(output_dir, "figures")
state_dir <- file.path(output_dir, "states")
annotation_dir <- file.path(output_dir, "annotations")
dir.create(output_dir, recursive = TRUE, showWarnings = FALSE)
dir.create(fig_dir, recursive = TRUE, showWarnings = FALSE)
dir.create(state_dir, recursive = TRUE, showWarnings = FALSE)
dir.create(annotation_dir, recursive = TRUE, showWarnings = FALSE)

min_features <- 200
max_features <- 6000
max_counts <- Inf
max_percent_mt <- 20
min_cells_per_gene <- 3
n_hvg <- 3000
dims_use <- NULL
npcs <- 40
vars_to_regress <- c("nFeature_RNA", "nCount_RNA", "percent.mt")
cluster_resolutions <- c(0.1, 0.5)
mt_pattern <- "^MT-"
contamination_method <- "auto"
run_doubletfinder <- TRUE
doubletfinder_pN <- 0.25
doubletfinder_npcs <- 30
doubletfinder_resolution <- 0.1
axel_proxy <- "http://www.cirno999.cn:12306"
axel_connections <- 10

# SingleR 上游结果可选，复制其中一个即可：
# "merged_no_correction"
# "integrated_rpca"
# "integrated_harmony"
# "integrated_bbknn"
singleR_annotation_input_key <- "integrated_harmony"
singleR_label_field <- "label.main"
singleR_reference <- NULL

state_paths <- list(
  object_list_raw = file.path(state_dir, paste0(project_id, "_01_object_list_raw.qs")),
  object_list_contam = file.path(state_dir, paste0(project_id, "_02_object_list_contam.qs")),
  obj_qc = file.path(state_dir, paste0(project_id, "_03_obj_qc.qs")),
  obj_singlet_merged_raw = file.path(state_dir, paste0(project_id, "_04_obj_singlet_merged_raw.qs")),
  obj_singlet_merged = file.path(state_dir, paste0(project_id, "_05_obj_singlet_merged.qs")),
  merged_no_correction = file.path(state_dir, paste0(project_id, "_06_merged_no_correction.qs")),
  integrated_rpca = file.path(state_dir, paste0(project_id, "_07_integrated_rpca.qs")),
  integrated_harmony = file.path(state_dir, paste0(project_id, "_08_integrated_harmony.qs")),
  integrated_bbknn = file.path(state_dir, paste0(project_id, "_09_integrated_bbknn.qs")),
  seurat_results = file.path(state_dir, paste0(project_id, "_10_seurat_results.qs")),
  session_info = file.path(state_dir, paste0(project_id, "_11_session_info.txt")),
  annotation_dir = annotation_dir
)

multiplet_rates_10x <- data.frame(
  Multiplet_rate = c(0.004, 0.008, 0.0160, 0.023, 0.031, 0.039, 0.046, 0.054, 0.061, 0.069, 0.076),
  Loaded_cells = c(800, 1600, 3200, 4800, 6400, 8000, 9600, 11200, 12800, 14400, 16000),
  Recovered_cells = c(500, 1000, 2000, 3000, 4000, 5000, 6000, 7000, 8000, 9000, 10000)
)

cleanup_vars <- function(var_names, envir = .GlobalEnv) {
  existing <- intersect(var_names, ls(envir = envir, all.names = TRUE))
  if (length(existing) > 0) {
    rm(list = existing, envir = envir)
  }
  invisible(gc())
}

qread_or_stop <- function(path, label = basename(path)) {
  if (!file.exists(path)) {
    stop(sprintf("Required upstream file '%s' was not found: %s", label, path), call. = FALSE)
  }
  qread(path)
}

download_with_axel <- function(url, dest_path) {
  dir.create(dirname(dest_path), recursive = TRUE, showWarnings = FALSE)
  status <- system2(
    "axel",
    args = c("-n", as.character(axel_connections), "-o", dest_path, url),
    env = c(
      sprintf("http_proxy=%s", axel_proxy),
      sprintf("https_proxy=%s", axel_proxy),
      sprintf("all_proxy=%s", axel_proxy)
    )
  )
  if (!identical(status, 0L) || !file.exists(dest_path)) {
    stop(sprintf("axel download failed for: %s", url), call. = FALSE)
  }
  dest_path
}

calculate_min_pc <- function(obj, reduction = "harmony") {
  stdv <- obj[[reduction]]@stdev
  percent_stdv <- (stdv / sum(stdv)) * 100
  cumulative <- cumsum(percent_stdv)
  co1 <- which(cumulative > 90 & percent_stdv < 5)[1]
  co2 <- sort(
    which((percent_stdv[1:(length(percent_stdv) - 1)] -
             percent_stdv[2:length(percent_stdv)]) > 0.1),
    decreasing = TRUE
  )[1] + 1
  min_pc <- min(co1, co2, na.rm = TRUE)
  if (!is.finite(min_pc)) {
    min_pc <- min(30, length(stdv))
  }
  max(2, min_pc)
}

resolve_dims_use <- function(obj, reduction = "harmony") {
  if (!is.null(dims_use)) {
    return(dims_use)
  }
  seq_len(calculate_min_pc(obj, reduction = reduction))
}

preprocess_standard <- function(obj, npcs_use = npcs) {
  obj <- NormalizeData(
    obj,
    normalization.method = "LogNormalize",
    scale.factor = 10000,
    margin = 1,
    assay = "RNA"
  )
  obj <- FindVariableFeatures(obj, nfeatures = n_hvg)
  obj <- ScaleData(obj, vars.to.regress = vars_to_regress)
  obj <- RunPCA(obj, features = VariableFeatures(obj), npcs = npcs_use)
  obj
}

estimate_multiplet_rate <- function(cell_number) {
  idx <- which(multiplet_rates_10x$Recovered_cells < cell_number)
  if (length(idx) == 0) {
    return(multiplet_rates_10x$Multiplet_rate[1])
  }
  multiplet_rates_10x$Multiplet_rate[max(idx)]
}

get_singleR_cluster_column <- function(input_key) {
  switch(
    input_key,
    merged_no_correction = "RNA_snn_res.0.1",
    integrated_rpca = "RNA_snn_res.0.1",
    integrated_harmony = "RNA_snn_res.0.1",
    integrated_bbknn = "bbknn_res.0.1",
    stop(sprintf("Unsupported singleR_annotation_input_key: %s", input_key), call. = FALSE)
  )
}

build_singleR_output_paths <- function(input_key) {
  list(
    table_csv = file.path(annotation_dir, paste0(project_id, "_", input_key, "_SingleR_res0.1.csv")),
    annotated_qs = file.path(annotation_dir, paste0(project_id, "_", input_key, "_SingleR_annotated.qs"))
  )
}

run_doubletfinder_one_sample <- function(o) {
  o <- NormalizeData(
    o,
    normalization.method = "LogNormalize",
    scale.factor = 10000,
    margin = 1,
    assay = "RNA"
  )
  o <- FindVariableFeatures(o, nfeatures = n_hvg)
  o <- ScaleData(o, vars.to.regress = vars_to_regress)
  o <- RunPCA(o, features = VariableFeatures(o), npcs = doubletfinder_npcs)
  min_pc <- calculate_min_pc(o, reduction = "pca")
  o <- RunUMAP(o, reduction = "pca", dims = 1:min_pc, reduction.name = "umap")
  o <- FindNeighbors(o, reduction = "pca", dims = 1:min_pc)
  o <- FindClusters(o, resolution = doubletfinder_resolution)
  sweep.res.hcc <- DoubletFinder::paramSweep(o, PCs = 1:min_pc, sct = FALSE)
  sweep.stats <- DoubletFinder::summarizeSweep(sweep.res.hcc, GT = FALSE)
  bcmvn <- DoubletFinder::find.pK(sweep.stats)
  pK_bcmvn <- as.numeric(as.vector(bcmvn$pK[which.max(bcmvn$BCmetric)]))
  homotypic.prop <- DoubletFinder::modelHomotypic(o$seurat_clusters)
  multiplet_rate <- estimate_multiplet_rate(nrow(o@meta.data))
  nExp.poi <- round(multiplet_rate * nrow(o@meta.data))
  nExp.poi.adj <- round(nExp.poi * (1 - homotypic.prop))
  DoubletFinder::doubletFinder(
    seu = o,
    PCs = 1:min_pc,
    pN = doubletfinder_pN,
    pK = pK_bcmvn,
    nExp = nExp.poi.adj,
    sct = FALSE
  )
}

remove_doublets_by_sample <- function(obj) {
  sample_obj <- SplitObject(obj, split.by = "sample")
  h <- lapply(names(sample_obj), function(nm) run_doubletfinder_one_sample(sample_obj[[nm]]))
  names(h) <- names(sample_obj)
  h.singlet <- lapply(names(sample_obj), function(nm) {
    o <- h[[nm]]
    colnames(o@meta.data)[grepl("DF.classifications", colnames(o@meta.data))] <- "doublet_finder"
    subset(o, subset = doublet_finder == "Singlet")
  })
  names(h.singlet) <- names(sample_obj)
  if (length(h.singlet) == 1) {
    singlet_merged <- h.singlet[[1]]
  } else {
    singlet_merged <- merge(h.singlet[[1]], y = h.singlet[-1], add.cell.ids = names(h.singlet))
  }
  singlet_merged <- JoinLayers(singlet_merged)
  cleanup_vars(c("h", "h.singlet", "sample_obj"), envir = environment())
  singlet_merged
}

run_contamination_correction <- function(obj) {
  # 如果原始矩阵和 filtered 矩阵都存在，优先在这里补充 SoupX 逻辑。
  # 如果只有 filtered 矩阵，可以在这里补充 DecontX 逻辑。
  obj
}

qc_filter_one <- function(obj) {
  obj[["percent.mt"]] <- PercentageFeatureSet(obj, pattern = mt_pattern)
  subset(
    obj,
    subset = nFeature_RNA >= min_features &
      nFeature_RNA <= max_features &
      nCount_RNA <= max_counts &
      percent.mt <= max_percent_mt
  )
}

collect_seurat_results <- function() {
  result_files <- c(
    merged_no_correction = state_paths$merged_no_correction,
    integrated_rpca = state_paths$integrated_rpca,
    integrated_harmony = state_paths$integrated_harmony,
    integrated_bbknn = state_paths$integrated_bbknn
  )
  out <- lapply(names(result_files), function(nm) {
    if (file.exists(result_files[[nm]])) {
      qread(result_files[[nm]])
    } else {
      NULL
    }
  })
  names(out) <- names(result_files)
  Filter(Negate(is.null), out)
}
```

## 02. Raw Input Processing

```{r raw-input-processing}
# 只在这一段从原始文件开始处理；后续模块都依赖 qread/qsave 串联状态。
read_one_sample <- function(sample_id, path) {
  counts <- Read10X(data.dir = path)
  one_meta <- sample_meta[match(sample_id, sample_meta$sample_id), , drop = FALSE]
  obj <- CreateSeuratObject(
    counts = counts,
    project = project_id,
    min.cells = min_cells_per_gene,
    meta.data = one_meta
  )
  obj$sample_id <- sample_id
  obj$sample <- sample_id
  obj <- RenameCells(obj, add.cell.id = sample_id)
  obj
}

# sample_paths <- c("<SAMPLE_1_PATH>", "<SAMPLE_2_PATH>")
# names(sample_paths) <- sample_meta$sample_id
# object_list <- setNames(lapply(names(sample_paths), function(x) read_one_sample(x, sample_paths[[x]])), names(sample_paths))
# qsave(object_list, state_paths$object_list_raw)
# cleanup_vars(c("object_list", "sample_paths"))
```

## 03. RNA Contamination Correction

```{r contamination-correction}
# object_list <- qread_or_stop(state_paths$object_list_raw, "object_list_raw")
# object_list <- lapply(object_list, run_contamination_correction)
# qsave(object_list, state_paths$object_list_contam)
# cleanup_vars(c("object_list"))
```

## 04. QC Filtering

```{r qc-filtering}
# object_list <- qread_or_stop(state_paths$object_list_contam, "object_list_contam")
# qc_list <- lapply(object_list, qc_filter_one)
# obj_qc <- if (length(qc_list) == 1) qc_list[[1]] else merge(qc_list[[1]], y = qc_list[-1], add.cell.ids = names(qc_list))
# obj_qc <- JoinLayers(obj_qc)
# qsave(obj_qc, state_paths$obj_qc)
# cleanup_vars(c("object_list", "qc_list", "obj_qc"))
```

## 05. DoubletFinder Removal

```{r doubletfinder-removal}
# 按 sample 拆分后分别去双，再把 singlet 重新合并。
# obj_qc <- qread_or_stop(state_paths$obj_qc, "obj_qc")
# obj_singlet_merged_raw <- if (run_doubletfinder) remove_doublets_by_sample(obj_qc) else obj_qc
# qsave(obj_singlet_merged_raw, state_paths$obj_singlet_merged_raw)
# cleanup_vars(c("obj_qc", "obj_singlet_merged_raw"))
```

## 06. Re-preprocessing of Merged Singlets

```{r re-preprocessing-merged-singlets}
# 对去双后的合并对象重新跑一次标准流程，后续各整合模块都从这里继续分叉。
# obj_singlet_merged_raw <- qread_or_stop(state_paths$obj_singlet_merged_raw, "obj_singlet_merged_raw")
# obj_singlet_merged <- preprocess_standard(obj_singlet_merged_raw)
# qsave(obj_singlet_merged, state_paths$obj_singlet_merged)
# cleanup_vars(c("obj_singlet_merged_raw", "obj_singlet_merged"))
```

## 07. Merge-only Baseline

```{r merge-only-baseline}
# 这一支只用合并后的对象直接聚类，不做额外整合。
# obj_merged_no_correction <- qread_or_stop(state_paths$obj_singlet_merged, "obj_singlet_merged")
# dims_pca <- resolve_dims_use(obj_merged_no_correction, reduction = "pca")
# obj_merged_no_correction <- RunUMAP(obj_merged_no_correction, reduction = "pca", dims = dims_pca, reduction.name = "umap")
# obj_merged_no_correction <- FindNeighbors(obj_merged_no_correction, reduction = "pca", dims = dims_pca)
# obj_merged_no_correction <- FindClusters(obj_merged_no_correction, resolution = cluster_resolutions)
# qsave(obj_merged_no_correction, state_paths$merged_no_correction)
# cleanup_vars(c("obj_merged_no_correction", "dims_pca"))
```

## 08. RPCA Integration

```{r rpca-integration}
# obj_rpca <- qread_or_stop(state_paths$obj_singlet_merged_raw, "obj_singlet_merged_raw")
# obj_rpca[["RNA"]] <- split(obj_rpca[["RNA"]], f = obj_rpca$sample_id)
# obj_rpca$sample <- obj_rpca$sample_id
# obj_rpca <- preprocess_standard(obj_rpca)
# obj_rpca <- IntegrateLayers(
#   object = obj_rpca,
#   method = RPCAIntegration,
#   orig.reduction = "pca",
#   new.reduction = "integrated.rpca",
#   verbose = FALSE
# )
# obj_rpca <- JoinLayers(obj_rpca)
# dims_rpca <- resolve_dims_use(obj_rpca, reduction = "integrated.rpca")
# obj_rpca <- FindNeighbors(obj_rpca, reduction = "integrated.rpca", dims = dims_rpca)
# obj_rpca <- RunUMAP(obj_rpca, reduction = "integrated.rpca", dims = dims_rpca, reduction.name = "umap")
# obj_rpca <- FindClusters(obj_rpca, resolution = cluster_resolutions)
# qsave(obj_rpca, state_paths$integrated_rpca)
# cleanup_vars(c("obj_rpca", "dims_rpca"))
```

## 09. Harmony Integration

```{r harmony-integration}
# require_or_stop("harmony")
# obj_harmony <- qread_or_stop(state_paths$obj_singlet_merged_raw, "obj_singlet_merged_raw")
# obj_harmony$sample <- obj_harmony$sample_id
# obj_harmony <- preprocess_standard(obj_harmony)
# RunHarmony <- harmony::RunHarmony
# obj_harmony <- RunHarmony(obj_harmony, "sample")
# dims_harmony <- resolve_dims_use(obj_harmony, reduction = "harmony")
# obj_harmony <- FindNeighbors(obj_harmony, reduction = "harmony", dims = dims_harmony)
# obj_harmony <- RunUMAP(obj_harmony, reduction = "harmony", dims = dims_harmony, reduction.name = "umap")
# obj_harmony <- FindClusters(obj_harmony, resolution = cluster_resolutions)
# qsave(obj_harmony, state_paths$integrated_harmony)
# cleanup_vars(c("obj_harmony", "dims_harmony"))
```

## 10. BBKNN Integration

```{r bbknn-integration}
# require_or_stop("bbknnR")
# obj_bbknn <- qread_or_stop(state_paths$obj_singlet_merged_raw, "obj_singlet_merged_raw")
# obj_bbknn$sample <- obj_bbknn$sample_id
# obj_bbknn <- preprocess_standard(obj_bbknn)
# dims_bbknn <- if (is.null(dims_use)) calculate_min_pc(obj_bbknn, reduction = "pca") else max(dims_use)
# obj_bbknn <- bbknnR::RunBBKNN(
#   object = obj_bbknn,
#   batch_key = "sample",
#   reduction = "pca",
#   n_pcs = dims_bbknn,
#   graph_name = "bbknn",
#   run_TSNE = FALSE,
#   run_UMAP = TRUE,
#   UMAP_name = "umap",
#   seed = seed,
#   verbose = FALSE
# )
# obj_bbknn <- FindClusters(obj_bbknn, graph.name = "bbknn", resolution = cluster_resolutions)
# qsave(obj_bbknn, state_paths$integrated_bbknn)
# cleanup_vars(c("obj_bbknn", "dims_bbknn"))
```

## 11. SingleR Initial Annotation

```{r singler-initial-annotation}
# 默认针对 res0.1 的聚类结果做初步自动注释，并把结果保存到 annotations 目录。
# require_or_stop("SingleR")
# require_or_stop("SingleCellExperiment")
# if (is.null(singleR_reference)) {
#   stop("Set singleR_reference to a valid reference object before running the SingleR module.", call. = FALSE)
# }
# if (!singleR_label_field %in% colnames(SummarizedExperiment::colData(singleR_reference))) {
#   stop(sprintf("singleR_label_field '%s' was not found in singleR_reference.", singleR_label_field), call. = FALSE)
# }
# singleR_input_path <- state_paths[[singleR_annotation_input_key]]
# if (is.null(singleR_input_path)) {
#   stop(sprintf("Unknown singleR_annotation_input_key: %s", singleR_annotation_input_key), call. = FALSE)
# }
# obj_singleR <- qread_or_stop(singleR_input_path, singleR_annotation_input_key)
# cluster_col <- get_singleR_cluster_column(singleR_annotation_input_key)
# if (!cluster_col %in% colnames(obj_singleR@meta.data)) {
#   stop(sprintf("Cluster column '%s' was not found in the selected object.", cluster_col), call. = FALSE)
# }
# cluster_ids <- as.character(obj_singleR[[cluster_col]][, 1])
# sce_singleR <- as.SingleCellExperiment(obj_singleR, assay = "RNA")
# pred_singleR <- SingleR::SingleR(
#   test = sce_singleR,
#   ref = singleR_reference,
#   labels = SummarizedExperiment::colData(singleR_reference)[[singleR_label_field]],
#   clusters = cluster_ids
# )
# pred_singleR_df <- as.data.frame(pred_singleR)
# pred_singleR_df$cluster_id <- rownames(pred_singleR_df)
# pred_singleR_df <- pred_singleR_df[, c("cluster_id", setdiff(colnames(pred_singleR_df), "cluster_id")), drop = FALSE]
# singleR_labels <- pred_singleR_df$pruned.labels
# if (all(is.na(singleR_labels))) {
#   singleR_labels <- pred_singleR_df$labels
# }
# names(singleR_labels) <- pred_singleR_df$cluster_id
# obj_singleR$singleR_res0.1_cluster <- cluster_ids
# obj_singleR$singleR_res0.1_label <- unname(singleR_labels[cluster_ids])
# singleR_outputs <- build_singleR_output_paths(singleR_annotation_input_key)
# data.table::fwrite(pred_singleR_df, file = singleR_outputs$table_csv)
# qsave(obj_singleR, singleR_outputs$annotated_qs)
# cleanup_vars(c("obj_singleR", "sce_singleR", "pred_singleR", "pred_singleR_df", "singleR_labels", "cluster_ids", "cluster_col", "singleR_outputs", "singleR_input_path"))
```

## 12. Optional Python Placeholder

```{python, eval=FALSE}
# Add optional Python code here when needed.
# Example: helper scripts for file preprocessing or external QC.
print("python chunk ready")
```

## 13. Final Results Assembly

```{r final-results-assembly}
# seurat_results <- collect_seurat_results()
# if (length(seurat_results) == 0) {
#   stop("No downstream result files were found. Run at least one baseline/integration module first.", call. = FALSE)
# }
# qsave(seurat_results, state_paths$seurat_results)
# cleanup_vars(c("seurat_results"))
```

## 14. Diagnostic Plots

```{r diagnostic-plots}
# seurat_results <- qread_or_stop(state_paths$seurat_results, "seurat_results")
# for (nm in names(seurat_results)) {
#   obj_plot <- seurat_results[[nm]]
#   p <- DimPlot(obj_plot, reduction = "umap", group.by = "sample", pt.size = 0.3)
#   ggsave(file.path(fig_dir, paste0(project_id, "_", nm, "_sample_umap.png")), plot = p, width = 8, height = 6)
#   cleanup_vars(c("obj_plot", "p"))
# }
# cleanup_vars(c("seurat_results"))
```

## 15. Final Validation

```{r final-validation}
# seurat_results <- qread_or_stop(state_paths$seurat_results, "seurat_results")
# print(seurat_results)
# stopifnot(length(seurat_results) >= 1)
# stopifnot(all(vapply(seurat_results, inherits, logical(1), what = "Seurat")))
# stopifnot(all(vapply(seurat_results, function(x) "umap" %in% names(x@reductions), logical(1))))
# stopifnot(file.exists(state_paths$seurat_results))
# writeLines(capture.output(sessionInfo()), con = state_paths$session_info)
# cleanup_vars(c("seurat_results"))
```
